---
title: '"MTCARS" Statistical Analysis'
author: "Dan Mailman"
date: last-modified
filters:
  - include-code-files
header-includes:
  - \usepackage{float}                            # Control figure, table placement 
  - \floatplacement{table}{H}                     # 'H' (here): ensure tables stay in place
  - \floatplacement{figure}{H}                    # 'H' (here): ensure tables stay in place
  - \usepackage{fontspec}                         # Allow font customization
  - \setmainfont{Times New Roman}                 # Set main font to Times New Roman
  - \setsansfont{Arial}                           # Set sans-serif font to Arial
  - \setmonofont[Scale=0.7]{Courier New}          # Set  monospace font to Scaled Courier New
  - \RedeclareSectionCommand[beforeskip=-1pt, afterskip=1pt]{section}
  - \RedeclareSectionCommand[beforeskip=-1pt, afterskip=1pt]{subsection}
  - \RedeclareSectionCommand[beforeskip=-1pt, afterskip=1pt]{subsubsection}
  - \BeforeTOCHead[toc]{\vspace*{-5pt}}
  - \AfterTOCHead[toc]{\vspace*{5pt}}
  - \usepackage{tocloft}                          # TOC formatting
  - \setlength{\cftbeforetoctitleskip}{0pt}       # Space before TOC title
  - \setlength{\cftaftertoctitleskip}{5pt}        # Space after TOC title
  - \setlength{\cftparskip}{0pt}                  # Paragraph skip in TOC
  - \setlength{\cftbeforesecskip}{2pt}            # Space between TOC 
  - \usepackage{geometry}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{booktabs}
  - \usepackage{listings}
  - \usepackage{xcolor}
  - \usepackage{color}
  - \usepackage{amsmath}

format:
  pdf:
    toc: true
    toc_depth: 1
    toc-title: Contents
    number-sections: true
    number-depth: 3
    highlight-style: Radical
    geometry:
      - top=.75in
      - left=.75in
      - bottom=.75in
      - right=.75in
    fontsize: 10pt
    colorlinks: true
    papersize: letter
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Define usable dimensions
inMargin  =  0.75
inPgW     =  8.5
inPgH     = 11.0
inPrtW    = inPgW - (2 * inMargin)
inPrtH    = inPgH - (2 * inMargin)
pacman::p_load(knitr,                   
               tidyverse,
               tinytex,
               kableExtra,   # kable_styling
               grid)         # grid.draw
options(scipen = 999)
```

```{r RunCode, warning=FALSE, include=FALSE}
source("mtcars_analysis.R")
```

\pagenumbering{gobble} <!-- Turns off page numbering for the first page -->
\listoftables
\listoffigures
\newpage
\pagenumbering{arabic} <!-- Restarts page numbering from here -->

# Introduction

## Background

One piece of information that influences car-buyer choices is **miles per gallon** (**mpg**).  So, it would be useful to be able to predict **mpg** from other car-model variables.

the **mtcars** dataset provided with r is an example dataset for variables and **mpg** data.

## Objectives

This analysis attempts to find a reasonably accurate model using **mtcars** to predict the **dependent variable** (**DV**), **mpg** from available **independent variables** (**IV**s).

@fig-DistMPG shows the distribution of **mpg**.


The goal is to:

- Distinguish among significant and insignificant IVs.
- Rank significance for IVs.
- Produce a Model/Formula for predicting **mpg** from significant IVs.

# Data Famililiarization

Learning what each of the variables represents enables a preliminary classification that eliminates two variables from consideration as predictors for **mpg**. @tbl-VarClassTbl indicates **mtcars** IVs may be classified as **design** (specification for final mechanical assembly) and **measurement** (used to evaluate the finished assembly). By this criterion, **hp** and **qsec** are not best candidates for inclusion in predictive models for **mpg**.

To determine a preliminary ordering for model candidates, this study hypothesizes 

1. Measurement IVs are not likely to be the best candidates for final model IVs.
1. IVs highly correlated to DV are better candidates for final model IVs.

# Data Conditioning

To facilitate correlation (via spearman), IVs with low integral counts of unique values are treated as ordinals. IVs with larger variation or real values are treated as numerics.
@tbl-OrdsSummTbl and @tbl-NumsSummTbl summarize ordinal and numeric IVs.

# Model IV Candidate Ordering by Correlation

IV/DV correlation magnitudes (whether positive or negative) are used to order IV candidates for inclusion in the model.  @tbl-IVModelOrdTbl shows positive and negative correlations to DV in diminishing degree order. Taking absolute values of correlation coefficients produces an experimental candidate order:

\begin{center}
\textbf{`r vCorrOrder`}
\end{center}

Removing the **measurement** variables **hp** and **qsec**:

\begin{center}
\textbf{`r vCorrOrder01`}
\end{center}

@tbl-MtCarsTbl shows conditioned, re-ordered **mtcars** data.

@tbl-mCorrsTbl and @fig-CorrAllIVs show correlations with **mpg** and among IVs.
@tbl-mCorrsDepVarAndGoodIVsTbl and figure @fig-CorrAllIVs01 show the same information removing measurement IVs.

# Eliminating Multicollinearity and Forming a Preliminary Hypothesis

Highly correlated IVs are candidates for multicollinearity.  @tbl-CorrGrpsTbl shows highly correlated IV groups.  The IVs in the lower two groups don't correlate highly with **mpg**, so are not candidates for the predictive model.

At this point, we are able to hypothesize that the model will not include **hp** at all, and will only include a linear combination of **cyl**, **disp**, and **wt**.  A little research into these variables indicates that **disp** is to large degree a physical function of **cyl** and that **wt** is independent of **cyl**.  So, before using r tools to evaluate models, we can hypothesize that the model should be based on **cyl** and **wt**.

# Using `step()` to Select Predictors

Using `step()` with models based on no IVs and the full set of IVs results in a ***STEP_MODEL*** formula/model for optimal set of predictors:

$$\mathbf{mpg = 38.75179 - 3.16697 \cdot wt - 0.94162 \cdot cyl - 0.01804 \cdot hp}$$

Removing **hp** from ***STEP_MODEL*** for the ***WT_CYL_MODEL***:

$$\mathbf{mpg = 39.6863 - 3.1910 \cdot wt - 1.5078 \cdot cyl}$$

# Model Summary, Comparison, Conclusion

- **Model with wt and cyl only**
  - Coefficients: Both predictors (**wt** and **cyl**) are highly significant (p < 0.01).
  - Multiple R-squared: 0.8302
  - Adjusted R-squared: 0.8185
  - F-statistic: 70.91 on 2 and 29 DF, p-value: 6.809e-12
- **Model with wt, cyl, and hp**
  - Coefficients:
    - **wt** and **Intercept** remain highly significant.
    - **cyl** significance drops (p < 0.1, less significant than in the first model).
    - **hp** is not statistically significant (p > 0.1).
  - Multiple R-squared: 0.8431
  - Adjusted R-squared: 0.8263
  - F-statistic: 50.17 on 3 and 28 DF, p-value: 2.184e-11

The **wt/cyl** model is preferable due to its simplicity and the statistical significance of all included predictors. The addition of **hp** in the step_model does not provide enough benefit in terms of explained variance to justify its inclusion, especially given its lack of statistical significance and the minimal increase in adjusted R-squared. @tbl-TblComparePredictions and @fig-CompareModels summarize the comparison of the two models.  Results of programmatic comparison via this study's `compare_two_models()` (see appendix) follows.

```{r compare_models}
mdlWtCylHp = lm(mpg ~ wt + cyl + hp, data = mtcars)
mdlWtCyl   = lm(mpg ~ wt + cyl,      data = mtcars)
compare_two_models(mdlWtCylHp, mdlWtCyl)
```

\newpage

# Tables

```{r VarClassTbl, echo=FALSE}
#| label: tbl-VarClassTbl
#| tbl-cap: Variable Classes
kable(dfClassTbl,
      format = "latex", booktabs = TRUE, row.names = FALSE,  linesep = "",) %>% 
kable_styling(full_width = FALSE, font_size = 7, latex_options = c("scale_down"))
```

```{r OrdsSummTbl, fig.pos="H", echo=FALSE}
#| label: tbl-OrdsSummTbl
#| tbl-cap: Ordinal Variables Summary
kable(dfOrdsSummTbl,
      format = "latex", booktabs = TRUE, row.names = FALSE,  linesep = "",) %>% 
kable_styling(full_width = FALSE, font_size = 7, latex_options = c("scale_down"))
```

```{r NumsSummTbl, echo=FALSE}
#| label: tbl-NumsSummTbl
#| tbl-cap: Numeric Variables Summary
kable(dfNumsSummTbl,
     format = "latex", booktabs = TRUE, row.names = FALSE,  linesep = "",) %>% 
kable_styling(full_width = FALSE, font_size = 7, latex_options = c("scale_down"))
```

```{r IVModelOrdTbl, echo=FALSE}
#| label: tbl-IVModelOrdTbl
#| tbl-cap: Ordered Model Candidate IVs
kable(dfIVModelOrdTbl, format = "latex", booktabs = TRUE, row.names = FALSE, 
      linesep = "") %>% 
  kable_styling(full_width = FALSE, font_size = 7, latex_options = c("scale_down"))
```

```{r MtCarsTbl, echo=FALSE, message=TRUE, warning=TRUE}
#| label: tbl-MtCarsTbl
#| tbl-cap: Conditioned Cars Data
dfCars  %>%  
kable(format = "latex", booktabs = TRUE, row.names = FALSE) %>%  
  kable_styling(full_width = FALSE, font_size = 7,latex_options = c("scale_down"))
```

```{r mCorrsTbl, echo=FALSE}
#| label: tbl-mCorrsTbl
#| tbl-cap: Initial Correlation Matrix
kable(mCorrs, format = "latex", booktabs = TRUE, row.names = TRUE, digits = 4,
      linesep = "") %>% 
  kable_styling(full_width = FALSE, font_size = 7,latex_options = c("scale_down"))
```

```{r mCorrsDepVarAndGoodIVsTbl, echo=FALSE}
#| label: tbl-mCorrsDepVarAndGoodIVsTbl
#| tbl-cap: Correlation Matrix w/o hp, qsec
kable(mCorrsDepVarAndGoodIVs, format = "latex", booktabs = TRUE, row.names = TRUE, digits = 4,
      linesep = "") %>% 
  kable_styling(full_width = FALSE, font_size = 7,latex_options = c("scale_down"))
```

```{r CorrGrpsTbl, echo=FALSE}
#| label: tbl-CorrGrpsTbl
#| tbl-cap: Multicollinearity Analysis
kable(dfCorrGrpsTbl, format = "latex", booktabs = TRUE, row.names = FALSE, digits = 4,
      linesep = "") %>% 
  kable_styling(full_width = FALSE, font_size = 7,latex_options = c("scale_down"))
```

```{r TblComparePredictions,echo=FALSE}
#| label: tbl-TblComparePredictions
#| tbl-cap: Model Predictions Comparison
kable(comparison_df, format = "latex", booktabs = TRUE, row.names = FALSE, digits = 4,
      linesep = "") %>% 
  kable_styling(full_width = FALSE, font_size = 7,latex_options = c("scale_down"))
```

\newpage

# Figures

```{r DistMPG, echo=FALSE, fig.height=2, message=FALSE, warning=FALSE}
#| label: fig-DistMPG
#| fig-cap: MPG Distribution
GetDistribPlot(dfCars,'mpg')
```


```{r CorrAllIVs, echo=FALSE}
#| label: fig-CorrAllIVs
#| fig-cap: Correlations with MPG and Ordered IVs
arranged_plots = ArrangeAndResizeSquarePlots(vAllDepVarPlotRows, 
                                             n_rows = length(vCorrOrder), 
                                             n_cols = length(vCorrOrder))
grid.draw(arranged_plots)
```

```{r CorrAllIVs01, echo=FALSE}
#| label: fig-CorrAllIVs01
#| fig-cap: Ordered IV Correlations (w/o HP, QSEC)
arranged_plots = ArrangeAndResizeSquarePlots(vAllDepVarPlotRows01, 
                                             n_rows = length(vCorrOrder01), 
                                             n_cols = length(vCorrOrder01))
grid.draw(arranged_plots)
```

```{r CompareModels, echo=FALSE,fig.height=2, warning=FALSE, message=FALSE}
#| label: fig-CompareModels
#| fig-cap: Predictive Model Comparison
ggplot(comparison_df, aes(x = Actual)) +
  geom_smooth(method = "lm", aes(y = mdlWtCyl, color = "mdlWtCyl"), se = FALSE) +
  geom_smooth(method = "lm", aes(y = mdlWtCylHp, color = "mdlWtCylHp"), se = FALSE) +
  geom_line(aes(y = Actual, color = "Actual")) +
  labs(x = "Actual MPG", y = "Predicted MPG") +
  scale_color_manual(values = c("mdlWtCylHp" = "red",
                                "mdlWtCyl" = "blue",
                                "Actual" = "black")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Legend"))
```

\newpage

# Appendix - Code

```{.r include="mtcars_analysis.R"}
```
